Formalizing Commutative Algebra in Coq: \\ Nakayama's Lemma\thanks{Source code for this work is available on the following site: \url{https://github.com/ku-sldg/algebra}}

### 2023-01-04 nakayama/paper.tex
Exposition updates, mostly on section on Axiom of Choice

\author{ 
	Andrew Cousino \\ {\ttfamily acousino@ku.edu}
	\and 
	Emily E.\ Witt\thanks{Witt acknowledges support from NSF CAREER Award
    DMS-1945611 and the 2022-23 Ruth I.\, Michler Memorial Prize from the
    Association for Women in Mathematics.} \\ {\ttfamily witt@ku.edu}
  \and 
	Perry Alexander \\ {\ttfamily palexand@ku.edu}
}

@@

\noindent \textbf{Keywords:}
Formalization of Mathematics,
Formal Proof,
Commutative Algebra,
Commutative Ring,
Local Ring,
Ideal,
Module over a Ring,
Finitely Generated Module.

@@

The notion of a module over a ring is an extension of the linear-algebraic
notion of a vector space over a field, ubiquitous in mathematics and its
applications. Less frequently referred to as the \emph{Krull-Azumaya theorem}
\footnote{Hideyuki Matsumura explains in his book \emph{Commutative Algebra}
  \cite{matsumura}: ``{This simple but important lemma is due to T.\ Nakayama,
  G.\ Azumaya, and W.\ Krull. Priority is obscure, and although it is usually
  called the Lemma of Nakayama, late Prof.\  Nakayama did not like the
  name.''}}
\,\cite{nagata}, Nakayama's Lemma describes one way that a finitely generated
module over an arbitrary commutative ring acts like a vector space over a
field. True to the convention that ``lemma'' often refers to a result serving
as a stepping stone toward another goal, Nakayama's Lemma is applied widely
throughout the field, and the result is typically introduced in a first
graduate course in commutative algebra \cite{atiyah-macdonald, matsumura,
eisenbud}.

@@

\subsubsection{Commutative rings}
In abstract algebra, the quintessential example of a commutative ring is the
set of integers
\[\mathbb{Z} = \{ \ldots, -3, -2, -1, 0, 1, 2, 3, \ldots \}.\]
using the natural definitions of addition and multiplication.

@@

\subsubsection{Ideals of commutative rings}
The concept of an {ideal} of a ring can be thought of as an extension of the
notion of an integer $n$ in the ring of integers $\mathbb{Z}$. An \emph{ideal}
of commutative ring $R$ is a subset $I$ of $R$ that is itself an abelian group
under addition, which also satisfies the following ``absorption'' property:
Given any element $a$ of $I$, the product  $x \cdot a$ is again in $I$ for any
ring element $x \in R$.

@@

\subsubsection{Local rings}
A commutative ring is \emph{local} if it has exactly one maximal ideal. 
Every field is local since the only proper ideal of a field is the zero ideal,
though by our observations above, the ring of integers is not local. 
However, the set of all rational numbers that can be written with an odd
denominator does form a subring of all rational numbers, and its unique maximal
ideal consists of the elements with even numerator; in fact, this ring is the
so-called \emph{localization} of $\mathbb{Z}$ at the maximal ideal
$2\mathbb{Z}$. The ring of integers modulo $n>1$ is local if and only if $n$ is
a power of a prime number $p$, in which case the unique maximal ideal consists
of all multiples of $p$.

@@

\subsubsection{Modules over commutative rings}
Let  $R$ be a commutative ring. 
A \emph{module over $R$}, or \emph{$R$-module}, is an abelian group $M$ under a
binary operation $+$, and a {scalar multiplication} $R \times M \to M$ denoted
$\cdot$, satisfying the following properties for all $r, s \in R$
and $u, v \in M$.

@@

\section{Constructing the Formal Proof}
Beyond formalizing the relevant structures from higher algebra, we also
formally establish some basic theory from the field in Coq. For instance, 
consider the notion of a \emph{unit} of a
commutative ring $R$, which is an element $a \in R$ with a multiplicative inverse,
i.e., there exists an element $a^{-1} \in R$ for which $x^{-1} \cdot x$ is the
multiplicative identity $1 \in R$.  In fact, it $x$ is a unit, its inverse is unique.

\emily[inline]{Emily: I changed statement to \emph{if and only if} below; do we need to update formal proof?}
Suppose that $I$ is an ideal of a commutative rings $R$.  
We formally state prove the following lemma: If $I$ contains a
unit $a$ if and only if $I$ is the trivial ideal, i.e., $I=R$. 
The informal proof logic is as follows:  Due to the absorption property of $I$,  
$a^{-1}  \cdot a= 1 \in I$.  Hence for every element $r$ of $R$,
$r = r \cdot 1$ is also in $I$, i.e., $I=R$. 
On the other hand, if $I = R$, then the multiplicative identity $1$ is a unit in $I$.

@@

\begin{quote}
	Fix a non-unit $x$ in a commutative ring $R$, and  \(I_{1}\) to be the principal 
	ideal generated by $x$, i.e., the smallest ideal containing the element $x$,
    which consists of all $R$-multiples of it.  No element of $I_1$ is a unit, since if  
    $y \in R$ were the multiplicative inverse of some $R$-multiple
    $r \cdot x$, then $1 = y \cdot (r \cdot x) = (y\cdot r) \cdot x$, showing that $x^{-1} = y \cdot r$, 
    which contradicts the assumption that $x$ is not a unit. Hence by the lemma described earlier in this section, 
    $I_1$ is proper.

    If \(I_{1}\) is not a maximal ideal of $R$, then by definition there exists a strictly larger
    proper ideal \(I_{2}\) of $R$, i.e.,
    $x \in I_{1}\subsetneq I_{2} \subsetneq R$. Likewise, if  \(I_{2}\) is again not
    maximal, then there exists a strictly larger proper ideal
    $I_{3}$, which, if not maximal, is strictly contained in another ideal, $I_4$, etc.

  Continuing in this way, if  this process terminates, we have successfully identified a
    maximal ideal of $R$ that contains $x$, and if not, one obtains an infinite, strictly increasing chain of proper ideals of $R$, each containing $x$, 
    \[I_{1}\subsetneq I_{2}\subsetneq I_{3}\subsetneq\cdots\subsetneq I_n \subsetneq \cdots. \]
    It is straightforward to verify that an increasing union of ideals 
    is, by definition, an ideal containing every ideal in the union.  Moreover, if each ideal in the union
    contains an element $x$, then the union certainly 
       contains \(x\) as well.

      Hence the set $S$ of all ideals of $R$, partially ordered by inclusion, 
       satisfies the property that every chain in $S$ has an upper bound in $S$.
    Therefore Zorn's lemma ensures the existence of a maximal ideal of \(R\) containing \(x\).
\end{quote}

The above argument has potentially infinitely many steps. A proof with infinitely
many steps is a problem in Coq. To see this from the above argument, suppose we
wrote a proof tactic \mintinline{coq}{generate_larger_ideal} that allows us to
go from one non-maximal ideal \(I_{k}\) which contains \(x\) to another ideal
\(I_{k+1}\) which also contains \(x\). Provided our first ideal \(I_{1}\) is
not maximal, our first invocation of \mintinline{coq}{generate_larger_ideal}
will produce another ideal \(I_{2}\). Our new ideal \(I_{2}\) may also not be
maximal, requiring that we invoke \mintinline{coq}{generate_larger_ideal}
again. This can lead to an infinite number of calls to the tactic
\mintinline{coq}{generate_larger_ideal} which will occupy any machine's
processor(s) forever, and further, the infinite number of ideals in the proof
state will fill any machine's memory. After we have overwhelmed any system with
these infinite number of calls to a single proof tactic, we still have to
invoke more tactics to finish the proof with an application of Zorn's lemma,
which finally guarantees us a maximal ideal containing \(x\). To summarize, we
are forced to run the following proof script which will never get past calling
\mintinline{coq}{generate_larger_ideal} infinitely many times.
\begin{minted}{coq}
  repeat generate_larger_ideal. (* must be called infinitely often *)
  zorns_lemma. (* called after the above non-terminating call finishes *)
\end{minted}
We chose to avoid this infinite proof script issue by including an axiom that
any non-unit \(x\) of a ring must be contained in some maximal ideal.

@@

\subsection{The Formal Statement of Nakayama's Lemma}
To start describing our proof, we formalize the statement of Nakayama's Lemma
in Coq.

@@

To use as a building block in the formal proof of Nakayama's Lemma, we first
formally stated and proved a lemma that gives a concrete description of the
elements of the submodule $\mathfrak{m} M$  appearing in the statement of the
theorem. This lemma applies more generally, to any finitely generated module
$M$ over a (not necessarily local) commutative ring, and any submodule of the
form $I M$, for $I$ an arbitrary ideal. By definition, $IM$ consists of scalar
combinations of elements of $M$ with coefficients in $I$. The lemma states that
the elements from $M$ appearing in the expression can be chosen to be from any
fixed finite generating set for $M$. In other words, given any generating set
$u_{1},\dots, u_{n} \in M$  of a  finitely generated $R$-module $M$, every
element $x$ of $I M$ can be written as a scalar combination $a_1 \cdot u_1 +
a_2  \cdot u_2 + \cdots + a_n  \cdot u_n$, where each $a_i \in I$.

@@

The proof follows from a straightforward argument based on definitions, by
induction on the number of elements in a fixed generating set for $M$, which we
informally describe: By definition, $x \in I M$ can  be written, for some
positive integer $k$ and elements $r_i \in R$ and $w_i \in M$,
$1 \leq i \leq k$, as $x = r_1 \cdot w_1 + r_2 \cdot w_2 + \cdots +
r_k \cdot w_k$. What's more, by definition of a finite generating set $u_{1},
\dots, u_{n}$ for $M$, each $w_i$ equals $c_{i1} \cdot u_1 + c_{i2} \cdot u_2 +
\cdots + c_{in} \cdot u_n$ for appropriate choices of $c_{ij} \in R$. Hence,
inductively applying associativity, 
\begin{align*}
  x = \sum_{i=1}^{k} r_i \cdot w_i
    = \sum_{i=1}^{k} r_i \cdot \left(\sum_{j=1}^n c_{ij} \cdot  u_j\right)
    = \sum_{j=1}^{n}\sum_{i=1}^{k} r_i \cdot (c_{ij} \cdot u_j) 
    = \sum_{j=1}^{n}\left(\sum_{i=1}^{k} c_{ij} \cdot r_i\right) \cdot u_j.   
\end{align*}
By absorption in ideals, each  $c_{ij}  \cdot r_i$ is in $I$, and since ideals
are closed under addition, we inductively conclude that the coefficient $a_j
:= c_{1j} \cdot  r_1 +  c_{2j} \cdot  r_2 +  \cdots c_{kj}\cdot r_k$ of $u_j$
is also in $I$.

Now we move forward to outline the formal proof of Nakayama's lemma. We proceed
by induction on the number of elements in a generating set for the $R$-module
$M$. The base case is the situation when \(M\) requires no generators, so $M$
consists solely of the empty scalar combination, i.e., the empty sum, which, by
convention, is the zero element. In other words, $M = 0$ by assumption, and
this is also precisely the conclusion of Nakayama's Lemma.  Hence the statement
trivially, without using the hypothesis that $M = \mathfrak{m} M$.

>>>>

\author{ 
	Andrew Cousino \\ {\ttfamily acousino@ku.edu}
	\and 
	Emily E.\ Witt\thanks{Witt acknowledges support from NSF CAREER Award
    DMS-1945611, the Keeler Intra-University Professorship from the University of Kansas, and the 2022-23 Ruth I.\, Michler Memorial Prize from the
    Association for Women in Mathematics.} \\ {\ttfamily witt@ku.edu}
  \and 
	Perry Alexander \\ {\ttfamily palexand@ku.edu}
}

@@

\noindent \textbf{Keywords:}
formalization of mathematics,
formal proof,
commutative algebra,
commutative ring,
local ring,
ideal,
module over a ring,
finitely generated module.

@@

The notion of a module over a ring is an extension of the linear-algebraic
notion of a vector space over a field, ubiquitous in mathematics and its
applications. Less frequently referred to as the \emph{Krull-Azumaya theorem}
\footnote{Hideyuki Matsumura explains in his text \emph{Commutative Algebra}
  \cite{matsumura}: ``{This simple but important lemma is due to T.\ Nakayama,
  G.\ Azumaya, and W.\ Krull. Priority is obscure, and although it is usually
  called the Lemma of Nakayama, late Prof.\  Nakayama did not like the
  name.''}}
\,\cite{nagata}, Nakayama's Lemma describes one way that a finitely generated
module over an arbitrary commutative ring acts like a vector space over a
field. True to the convention that ``lemma'' often refers to a result serving
as a stepping stone toward another goal, Nakayama's Lemma is applied widely
throughout the field, and the result is typically introduced in a first
graduate course in commutative algebra \cite{atiyah-macdonald, matsumura,
eisenbud}.

@@

\paragraph{Commutative rings}
In abstract algebra, the quintessential example of a commutative ring is the
set of integers
\[\mathbb{Z} = \{ \ldots, -3, -2, -1, 0, 1, 2, 3, \ldots \}.\]
using the natural definitions of addition and multiplication.

@@

\paragraph{Ideals of commutative rings}
The concept of an {ideal} of a ring can be thought of as an extension of the
notion of an integer $n$ in the ring of integers $\mathbb{Z}$. An \emph{ideal}
of commutative ring $R$ is a subset $I$ of $R$ that is itself an abelian group
under addition, which also satisfies the following ``absorption'' property:
Given any element $a$ of $I$, the product  $x \cdot a$ is again in $I$ for any
ring element $x \in R$.

@@

\paragraph{Local rings}
A commutative ring is \emph{local} if it has exactly one maximal ideal. 
Every field is local since the only proper ideal of a field is the zero ideal,
though by our observations above, the ring of integers is not local. 
However, the set of all rational numbers that can be written with an odd
denominator does form a subring of all rational numbers, and its unique maximal
ideal consists of the elements with even numerator; in fact, this ring is the
so-called \emph{localization} of $\mathbb{Z}$ at the maximal ideal
$2\mathbb{Z}$. The ring of integers modulo $n>1$ is local if and only if $n$ is
a power of a prime number $p$, in which case the unique maximal ideal consists
of all multiples of $p$.

@@

\paragraph{Modules over commutative rings}
Let  $R$ be a commutative ring. 
A \emph{module over $R$}, or \emph{$R$-module}, is an abelian group $M$ under a
binary operation $+$, and a {scalar multiplication} $R \times M \to M$ denoted
$\cdot$, satisfying the following properties for all $r, s \in R$
and $u, v \in M$.

@@

\section{Constructing the Formal Proof}
Not only to aid in the formalization of Nakayama's lemma, but also to

Beyond formalizing the relevant structures from higher algebra, we also
formally establish some basic theory from the field in Coq. For instance, 
consider the notion of a \emph{unit} of a
commutative ring $R$, which is an element $a \in R$ with a multiplicative inverse,
i.e., there exists an element $a^{-1} \in R$ for which $x^{-1} \cdot x$ is the
multiplicative identity $1 \in R$.  In fact, it $x$ is a unit, its inverse is unique.

\emily[inline]{Emily: Just changed statement to \emph{if and only if} below; we should check whether we should update its formal proof}
Suppose that $I$ is an ideal of a commutative rings $R$.  
We formally establish the following: If $I$ contains a
unit $a$ if and only if $I$ is the trivial ideal, i.e., $I=R$. 
The informal proof logic is as follows:  Due to the absorption property of $I$,  
$a^{-1}  \cdot a= 1 \in I$.  Hence for every element $r$ of $R$,
$r = r \cdot 1$ is also in $I$, i.e., $I=R$. 
On the other hand, if $I = R$, then the multiplicative identity $1$ is a unit in $I$.

@@

\begin{quote}
	Fix a non-unit $x$ in a commutative ring $R$, and  \(I_{1}\) to be the principal 
	ideal generated by $x$, i.e., the smallest ideal containing the element $x$,
    which consists of all $R$-multiples of it.  No element of $I_1$ is a unit, since if  
    $y \in R$ were the multiplicative inverse of some $R$-multiple
    $r \cdot x$, then $1 = y \cdot (r \cdot x) = (y\cdot r) \cdot x$, showing that $x^{-1} = y \cdot r$, 
    which contradicts the assumption that $x$ is not a unit. Hence by the statement described earlier in this section, 
    $I_1$ is proper.

   Either $I_1$ is a maximal ideal of $R$, or it is strictly contained in a maximal ideal, or there exists 
   an inifinite chain of strictly increasing
   proper ideals of the form 
    \begin{equation} \label{infinite-chain: e}
    I_{1}\subsetneq I_{2}\subsetneq I_{3}\subsetneq\cdots\subsetneq I_n \subsetneq \cdots. 
    \end{equation}  
   Indeed, suppose that $I_1$ is not maximal, and   
   assume that for some integer $n \geq 1$,  there exist proper, non-maximal ideals 
   $I_k$ of $R$, $1 < k \leq n$, such that for all such $k$,  $I_{k-1}$ is strictly contained in $I_k$.  
   Since $I_n$ is not maximal, there exists a proper ideal $I_{n+1}$ of $R$ such that $I_n \subsetneq I_{n+1}$.
      Then either $I_{n+1}$ is maximal, or we have extended the chain by one additional link.

%   
%    If \(I_{1}\) is not a maximal ideal of $R$, then by definition there exists a strictly larger
%    proper ideal \(I_{2}\) of $R$, i.e.,
%    $x \in I_{1}\subsetneq I_{2} \subsetneq R$. Likewise, if  \(I_{2}\) is again not
%    maximal, then there exists a strictly larger proper ideal
%    $I_{3}$, which, if not maximal, is strictly contained in another ideal, $I_4$, etc.
%

    It is straightforward to verify that an increasing union of ideals 
    is, by definition, an ideal containing every ideal in the union.  
Hence if, in this process, no maximal ideal is identified; i.e., we have ensured the existence of a chain of the form \eqref{infinite-chain: e},  then the increasing union of the $I_k$ is again an ideal, which contains $x$ since  $x \in I_1$.

%  Continuing in this way, if  this process terminates, we have successfully identified a
%    maximal ideal of $R$ that contains $x$, and if not, one obtains an infinite, strictly increasing chain of proper ideals of $R$, each containing $x$, 
%    \[I_{1}\subsetneq I_{2}\subsetneq I_{3}\subsetneq\cdots\subsetneq I_n \subsetneq \cdots. \]
%    It is straightforward to verify that an increasing union of ideals 
%    is, by definition, an ideal containing every ideal in the union.  Moreover, if each ideal in the union
%    contains an element $x$, then the union certainly 
%       contains \(x\) as well.

      Hence the set $X$ of all proper ideals of $R$ containing $x$, partially ordered by inclusion, 
       satisfies the property that every chain in $X$ has an upper bound in $X$.
    Therefore, Zorn's lemma ensures the existence of a maximal element of $X$, i.e., a maximal ideal of \(R\) containing \(x\).
\end{quote}

The above argument has potentially infinitely many steps, posing a Coq implementation
 problem. Indeed, suppose one were to write a proof tactic called \mintinline{coq}{generate_larger_ideal} 
allowing one to move from a non-maximal, proper ideal \(I\) that contains a fixed non-unit 
$x$ to a strictly larger ideal 
\(J \), which is again proper, and also contains \(x\).

After fixing a proper ideal $I_1$, provided that it is not maximal, 
our first invocation of \mintinline{coq}{generate_larger_ideal}
will produce another ideal, \(I_{2}\). It is possible that \(I_{2}\) is also non-maximal, 
requiring that we invoke \mintinline{coq}{generate_larger_ideal}
again, producing $I_3$, which again could be non-maximal, and so on, 
which could lead to an infinite number of calls to the tactic
\mintinline{coq}{generate_larger_ideal}
Of course, this would occupy any machine's
processor(s) forever, and further, the boundlessly increasing number of ideals in the proof
state would fill any machine's memory.

After we have overwhelmed any system with
these infinite number of calls to a single proof tactic, we still have to
invoke more tactics to finish the proof with an application of Zorn's lemma,
which finally guarantees us a maximal ideal containing \(x\). To summarize, we
are forced to run the following proof script which will never get past calling
\mintinline{coq}{generate_larger_ideal} infinitely many times.
\begin{minted}{coq}
  repeat generate_larger_ideal. (* must be called infinitely often *)
  zorns_lemma. (* called after the above non-terminating call finishes *)
\end{minted}
We chose to avoid this infinite proof script issue by including an axiom that
any non-unit \(x\) of a ring must be contained in some maximal ideal.

@@

\subsection{The Formal Statement}
We now present our formal statement of Nakayama's Lemma
in Coq, and describe a lemma called upon in our proof.

@@

To use as a building block in the formal proof of Nakayama's Lemma, we first
formally stated and proved a lemma that gives a concrete description of the
elements of the submodule $\mathfrak{m} M$  appearing in its statement. 
This lemma applies more generally, to an arbitrary finitely generated module
$M$ over a (not necessarily local) commutative ring, and any submodule of the
form $I M$, for $I$ an arbitrary ideal.

By definition, $IM$ consists of scalar
combinations of elements of $M$ with coefficients in $I$. The lemma states that
the elements from $M$ appearing in such an expression can be chosen to be from any
fixed finite generating set for $M$. In other words, given any generating set
$u_{1},\dots, u_{n} \in M$  for a  finitely generated $R$-module $M$, every
element $x$ of $I M$ can be written as a scalar combination $a_1 \cdot u_1 +
a_2  \cdot u_2 + \cdots + a_n  \cdot u_n$, where each $a_i \in I$.

@@

This lemma's proof follows from a straightforward argument based on definitions, by
induction on the number of elements in a fixed generating set for $M$, which we
informally describe: By definition, $x \in I M$ can  be written, for some
positive integer $k$ and elements $r_i \in R$ and $w_i \in M$,
$1 \leq i \leq k$, as $x = r_1 \cdot w_1 + r_2 \cdot w_2 + \cdots +
r_k \cdot w_k$. What's more, by definition of a finite generating set $u_{1},
\dots, u_{n}$ for $M$, each $w_i$ equals $c_{i1} \cdot u_1 + c_{i2} \cdot u_2 +
\cdots + c_{in} \cdot u_n$ for appropriate choices of $c_{ij} \in R$. Hence,
inductively applying associativity, 
\begin{align*}
  x = \sum_{i=1}^{k} r_i \cdot w_i
    = \sum_{i=1}^{k} r_i \cdot \left(\sum_{j=1}^n c_{ij} \cdot  u_j\right)
    = \sum_{j=1}^{n}\sum_{i=1}^{k} r_i \cdot (c_{ij} \cdot u_j) 
    = \sum_{j=1}^{n}\left(\sum_{i=1}^{k} c_{ij} \cdot r_i\right) \cdot u_j.   
\end{align*}
By absorption in ideals, each  $c_{ij}  \cdot r_i$ is in $I$, and since ideals
are closed under addition, we inductively conclude that the coefficient $a_j
:= c_{1j} \cdot  r_1 +  c_{2j} \cdot  r_2 +  \cdots c_{kj}\cdot r_k$ of $u_j$
is also in $I$.

\subsection{The Formal Proof}
We now outline the formal proof of Nakayama's lemma, and  proceed
by induction on the number of elements in a fixed generating set for the finitely generated $R$-module
$M$. The base case in this situation is when \(M\) requires no generators, so $M$
consists solely of the empty scalar combination, i.e., the empty sum, which, by
convention, is the zero element. In other words, $M = 0$ by assumption, and
this is also precisely the conclusion of Nakayama's Lemma.  Hence the statement holds
trivially, without using the hypothesis that $M = \mathfrak{m} M$.
